Lasciate ogne speranza, voi ch'intrate. (Canto III, line 9)

##################################################
3/30/26 Project Proposal
##################################################

This is a development log for the project to keep track of where I am in this project. Currently, I am finding data to work with for this ML project. 
However, it is first prudent to define the goal of this project before we get into the data collection. 

Goal: Find chemical motifs that are important and can perform chemical reactions of a certain type. For example, glycolic acid's motif is important for protection reactions. 
 - Structure: 
    - Model 1: 
        - A GNN using DFT data will find features for unknown molecules. 
        - A decision tree will then use those features for classification of the reaction type. 
    - Model 2: 
        - A LLM that will group certain atoms together as motifs and then use those motifs to classify the reaction type.
        - Analyze the motifs that the LLM finds and compare them to the features that the GNN finds.
    
 - Use and purpose: 
    - This model can be used to find motifs that are important for certain reactions. This can be used to design new molecules that have those motifs and can perform those reactions.
    - Doesn't do the chemical reaction prediction, but sure does make looking for organic reactions to work with theoretically easier.
    - KEY QUESTION: Does including the physics of DFT better than just looking at the motifs that the transformer finds?

Sources: 
    1. LLM and Organic Reactions learning not to do "alchemy": https://pmc.ncbi.nlm.nih.gov/articles/PMC6764164/ 
        - The Molecular Transformer demonstrates that chemical reactions can be predicted with high accuracy using sequence-based models, 
        but these models primarily learn statistical patterns in SMILES representations rather than underlying chemical mechanisms.
    2. GNN and how to represent molecules as graphs: https://arxiv.org/abs/1704.01212
        - Model molecules as graphs and iteratively propagate information between atoms, predicting quantum chemical properties.

##################################################
4/1/26 Checkpoint No. 1
##################################################

I need to have done "Data Preparation" for my models that I will be creating. Because SAVIO has been down for the past week, I have until tommorow night to get this done, 
but I don't think I need SAVIO just yet. We'll see. 

For the GNN, I need to find a dataset that has DFT data for molecules and their reactions. And for the decision tree, I need to find a dataset that has reaction types for molecules.
For the LLM, all I will need is the dataset that has the reaction types for the molecules. But I need to work backwards. 

The USPTO-50K dataset gives me what I need for reactions, and has 10 reaction types. It is drawn from the noisier USPTO dataset, with inumerous reaction labels. 
Before I can use this dataset, I need to know what elements are involved in the reactions, and then I can find DFT data for those molecules. I will write the python script. 

I have found a dataset that has DFT data for molecules (https://alchemy.tencent.com/) that has more elements than just CHONF. Remember to cite this when I write my paper, along w/ QM9. 
I will stick with the USPTO-50K dataset for the reaction data because there is literally no other dataset that has reaction types for the molecules.

So I just tried to push to GitHub but I realized that the sdf files from alchemy and others were too large to push. I have tried for the past few hours and I give up. 
I will just offload it to SAVIO or just keep it on my local machine for now. (Read .gitignore for more details.)

##################################################
4/2/26 Progress on the Code
##################################################

Using a fraction of the data so I can run this locally: 
    1. I need to now make the GNN and decision tree model. 
    2. I will also need to make the LLM model seperately.
    3. Compare how the two perform vs each other.

Let's start with just the GNN and make sure that works. I'm also gonna do 3-fold cross validation to make sure that the model is not overfitting.

##################################################
4/4/26 It came to me like a dream.
##################################################

Oh god, it's begun. I've started dreaming about this class. Well, more accurately, I woke up and the idea hit me. I don't need a decision tree, I just need to bake the GNN output values
and shove them into the transformer. That way, it would be a way better comparison of whether adding physics values would help me or not. All I have to do is figure out how to do a transformer
earlier than what was taught in class. 

    (1) As I have written the GNN, I started wondering if adding hydrogens would be important to the physics of all this? The smiles only has the elements, or maybe they're superfluous.
        They could also be terrible for run time. Maybe that's something I'll revisit at some point. 
    (2) My GNN does has like R^2 = 0.5 for small sample sizes which has me somewhat concerned. Something I should ask Juan. 

##################################################
4/6/26 Checkpoint No. 2
##################################################

I miss winter break. I miss SoCal. I also miss good food. I can not imagine returning to Berkeley for grad school unless I have a car.

aNyHOw, 'ML model construction and workflow development' check up time, and I think I will just focus on the skeleton of this ML model which is finishing making the GNN
and the transformer.

##################################################
4/6/26 - 4/9/26 Cleaning up the code
##################################################

I am going to use tutorial 11 and 12 to write the GNN and transformer over again to make sure I am doing it correctly. 
Write now, the first focus is making sure C and Cl do not get mixed up!

Surround the GNN layers around with MLP layers to increase the capacity because GNNS tend to have low capacity. 

##################################################
4/11/26 This Godforsaken GNN
##################################################

I am getting like R^2 of like 0.4 for the regresssion of the NN. I've swapped the the loss_type
to smoothl1 loss because these examples that it trains on can be very chemically different considering how many elements
there are.

I do wonder though, could it be features of the nodes that could be the problem? Like are they good tto help predict the homo, lumo, 
dG, etc.? I could also test it out. 

At the very least, I have made the pipeline to explore hyperparameters to find the best for the final GNN. 

Okay if this isn't working out, why don't I try something else? Let's try do a model that takes in the 3-d features 
of the molecule because that is way more powerful for quantum preditions in the alchemy dataset.

The GNN has the limitation that I can't do bond angles. Angles are defined by three nodes (ijk)
but the GINEConv only accepts features for node i. I can only do bond lengths. Hopefully that should help. 
If not I'll have to switch architechtures.

I mean to be fair, an bond lengths force a defined angle if nodes are connected, so maybe I 
can get away by adding the bond length only to improve R^2. 

IT WORKED! IM GONNA CRY UGHGHGHGHGHGUHUHJ. (Yes. I sound just like that in my head.)

Okay next fix I need to figure out is how to input the correct catgorical and continuosu features
into GINEConv! Atoms number (Carbon = 6) shouldn't go through the slop of linear algebra!
One hot encodding time! 

One hot coding make the gradient smoother and much cleaner. Now the next and final thing I want to consider before I burry this 
GNN as like 99.9\% done is the angle of the bond. 

The object of an angle is that it is a three node or 2 edge defined object 
(two rays make and angle or three points). How do I architechturely incorporates this into a GINEConv
model that only takes nodes and edges?

I went onto Gemini and asked and they said I could do this but that would mean switching to models that were much more computationally expensive. 
Also my R^2 seems to be scaling well, so it doesn't look like I really need it. 

Maybe, the most that I will do then is just take the average of the angle and put it into the node. 
The angle can be a node feature. The angle though will be like [120, 120, 120, 0, 0, 0]. Then another vector will be learn if there is a
bond there like [1, 1, 1, 0, 0, 0] which will be used to mask the angle since the model needs to discern angle vector. 
And the boolean vector which indicates if there is an angle or not.

If I have learned anything from today, if I have a problem, just turn it into a learnable. And let the linear algebra slop figure it out. 


